---
title: 'Age Old Battle: Fruits and Vegetables'
author: 'Alister Salmon'
date: 'February 10, 2024'
output: pdf_document
always_allow_html: true
---

```{r setup, message=FALSE, warning=FALSE}
  # clear the R environment
  rm(list=ls())

  # import required libraries
  library(dplyr)
  library(plotly)
  library(ggplot2)
  library(caret)
```


# Data Processing

```{r data_import}
  # import the data
  data.veg = data.frame(read.csv('Vegetable Prices 2020.csv'))
  data.fruit = data.frame(read.csv('Fruit Prices 2020.csv'))
  
```

```{r combining}
  # rename columns
  colnames(data.veg)[1] = 'Food'
  colnames(data.fruit)[1] = 'Food'
  
  # create type column
  data.veg['isFruit'] = 0
  data.fruit['isFruit'] = 1
  
  # combine them
  data = data.frame(rbind(data.veg, data.fruit))
  
  head(data)
  tail(data)
```


```{r export}
  # export the cleaned data
  write.csv(data, 'fruit_veg.csv', row.names = FALSE)
```


# Spilt Milk

Visualize the distribution of yield and price of fruits and vegetable by form

## Methodology

### Distribution Based on Yield

```{r dist_yield_fruit}
  # histogram
  ggplot(data.fruit, aes(x=Yield, fill=Form)) +
    geom_histogram() + ggtitle('Distribution of Fruit Yield') + ylab('Count')
  ggsave('visuals/fruit_hist_yield.jpg')
  
  # box plot of fresh
  ggplot(filter(data.fruit, Form=='Fresh'), aes(x=Yield)) + geom_boxplot() +
    theme(axis.text.y = element_blank(), axis.line.y = element_blank(),
          axis.ticks.y = element_blank()) +
    coord_fixed(ratio = 0.1)
  ggsave('visuals/fruit_boxplot_fresh.jpg')
```


```{r dist_yield_veg}
  # histogram
  ggplot(data.veg, aes(x=Yield, fill=Form)) +
    geom_histogram() + ggtitle('Distribution of Vegetable Yield') + ylab('Count')
  ggsave('visuals/veg_hist_yield.jpg')
  
  # box plot of fresh
  ggplot(filter(data.veg, Form=='Fresh'), aes(x=Yield)) + geom_boxplot() +
    theme(axis.text.y = element_blank(), axis.line.y = element_blank(),
          axis.ticks.y = element_blank())
    coord_fixed(ratio = 0.1)
  ggsave('visuals/veg_boxplot_fresh.jpg')
```

### Distribution Based on Price per Cup


```{r dist_price_fruit}
  # histogram
  ggplot(data.fruit, aes(x=CupEquivalentPrice, fill=Form)) +
    geom_histogram() + ggtitle('Distribution of Price per Cup of Fruit by From') +
    ylab('Count') + xlab('Price Per Cup')
  ggsave('visuals/fruit_hist_price.jpg')
  
  # box plot
  ggplot(data.fruit, aes(x=CupEquivalentPrice, y=Form)) + geom_boxplot() +
    xlab('Price per Cup') + ggtitle('Box Plots of Price per Cup of Fruit by Form')
  ggsave('visuals/fruit_boxplot_price.jpg')
```

```{r dist_price_veg}
  # histogram
  ggplot(data.veg, aes(x=CupEquivalentPrice, fill=Form)) +
    geom_histogram() + ggtitle('Distribution of Price per Cup of Vegetable by From') +
    ylab('Count') + xlab('Price per Cup')
  ggsave('visuals/veg_hist_price.jpg')
  
  # box plot
  ggplot(data.veg, aes(x=CupEquivalentPrice, y=Form)) + geom_boxplot() +
    xlab('Price per Cup')
  ggsave('visuals/veg_boxplot_price.jpg')
```

## Results

Price per cup of fruits and vegetables is roughly equivalent. Canned, dried, and frozen vegetables tend to be cheaper than that of fruit.



# Apples to Oranges

Compare fruits (prices and yield)

**Comparisons:**

Cheapest and most expensive fruits

Avg. price/yields of fresh vs canned vs juice vs dried vs frozen

## Methodology

```{r fruit_order_by_cup_price}
  # order the data by CupEquivalentPrice
  data.fruit.cup_price.ascending = 
    data.fruit[order(data.fruit$CupEquivalentPrice),]
  
  # get the least expensive
  head(select(data.fruit.cup_price.ascending, Food, Form, CupEquivalentPrice), 5)
  
  # get the most expensive
  tail(select(data.fruit.cup_price.ascending, Food, Form, CupEquivalentPrice), 5)
```

```{r fruit_order_by_yield}
  data.fruit.yield.ascending = data.fruit[order(data.fruit$Yield),]
  
  head(select(data.fruit.yield.ascending, Food, Form, Yield), 5)
  tail(select(data.fruit.yield.ascending, Food, Form, Yield), 5)
```



```{r fruit_avg_form_price}
  data.fruit.avg_form.price = aggregate(CupEquivalentPrice~Form, data=data.fruit, mean)
  
  ggplot(data=data.fruit.avg_form.price, aes(x=reorder(Form,-CupEquivalentPrice),
                                  y=CupEquivalentPrice)) + 
    geom_bar(stat='identity', fill='salmon') + theme_minimal() +
    xlab('Fruit Form') + ylab('Price per Cup') +
    geom_text(aes(label=sprintf('%.3f', CupEquivalentPrice)), vjust=1.6) +
    ggtitle('Average Price per Cup of Fruit by Form')
  ggsave('visuals/fruit_price_by_form.jpg')
```

```{r fruit_avg_form_yield}
  data.fruit.avg_form.yield = aggregate(Yield~Form, data=data.fruit, mean)

  ggplot(data=data.fruit.avg_form.yield, aes(x=reorder(Form,-Yield),
                                  y=Yield)) + 
    geom_bar(stat='identity', fill='salmon') + theme_minimal() +
    xlab('Fruit Form') + ylab('Yield per Cup') +
    geom_text(aes(label=sprintf('%.3f', Yield)), vjust=1.6) +
    ggtitle('Average Yield per Cup of Fruit by Form')
  ggsave('visuals/fruit_yield_by_form.jpg')
```

```{r fruit_price_per_yield}
  data.fruit.avg_form.avg_price_yield = 
    data.frame(data.fruit.avg_form.price$Form,
               data.fruit.avg_form.price$CupEquivalentPrice /
                 data.fruit.avg_form.yield$Yield)
  colnames(data.fruit.avg_form.avg_price_yield) = c('Form','Price')
  
  ggplot(data=data.fruit.avg_form.avg_price_yield, 
         aes(x=reorder(Form,-Price), y=Price)) + 
    geom_bar(stat='identity', fill='salmon') + theme_minimal() +
    xlab('Fruit Form') + ylab('Price') +
    geom_text(aes(label=sprintf('%.3f', Price)), vjust=1.6) +
    ggtitle('Average Price per Edible Cup of Fruit by Form')
  ggsave('visuals/fruit_price_per_yield.jpg')
```


## Results

Cheapest form of fruit by edible cup is juice. Fresh fruit is suprisingly cheap; however, this could be due to the bananas and watermelon which are abnormally cheap. The most expensive form is frozen. I did not expect this; however, these results do not take into account calorie count. If that data was available, I am sure that the results would be much different.


# Carrots to Cabbages

Compare vegetables (prices and yield)

**Comparisons:**

Cheapest and most expensive vegetables

Avg. price/yields of fresh vs canned vs frozen vs dried

## Methodology

```{r veg_order_by_cup_price}
  # order the data by CupEquivalentPrice
  data.veg.cup_price.ascending = 
    data.veg[order(data.veg$CupEquivalentPrice),]
  
  # get the least expensive
  head(select(data.veg.cup_price.ascending, Food, Form, CupEquivalentPrice), 5)
  
  # get the most expensive
  tail(select(data.veg.cup_price.ascending, Food, Form, CupEquivalentPrice), 5)
```

```{r veg_order_by_yield}
  data.veg.yield.ascending = data.veg[order(data.veg$Yield),]
  
  head(select(data.veg.yield.ascending, Food, Form, Yield), 5)
  tail(select(data.veg.yield.ascending, Food, Form, Yield), 5)
```



```{r veg_avg_form_price}
  data.veg.avg_form.price = aggregate(CupEquivalentPrice~Form, data=data.veg, mean)
  #data.veg.avg_form.price =
    #data.veg.avg_form.price[order(data.veg.avg_form.price$CupEquivalentPrice),]
  
  ggplot(data=data.veg.avg_form.price, aes(x=reorder(Form,-CupEquivalentPrice),
                                  y=CupEquivalentPrice)) + 
    geom_bar(stat='identity', fill='salmon') + theme_minimal() +
    xlab('Vegetable Form') + ylab('Price per Cup') +
    geom_text(aes(label=sprintf('%.3f', CupEquivalentPrice)), vjust=1.6) +
    ggtitle('Average Price per Cup of Vegetable by Form')
  ggsave('visuals/veg_price_by_form.jpg')
```

```{r veg_avg_form_yield}
  data.veg.avg_form.yield = aggregate(Yield~Form, data=data.veg, mean)
  #data.veg.avg_form.yield = 
    #data.veg.avg_form.yield[order(data.veg.avg_form.yield$Yield),]

  ggplot(data=data.veg.avg_form.yield, aes(x=reorder(Form,-Yield),
                                  y=Yield)) + 
    geom_bar(stat='identity', fill='salmon') + theme_minimal() +
    xlab('Vegetable Form') + ylab('Yield per Cup') +
    geom_text(aes(label=sprintf('%.3f', Yield)), vjust=1.6) +
    ggtitle('Average Yield per Cup of Vegetable by Form')
  ggsave('visuals/veg_yield_by_form.jpg')
```

```{r veg_price_per_yield}
  data.veg.avg_form.avg_price_yield = 
    data.frame(data.veg.avg_form.price$Form,
               data.veg.avg_form.price$CupEquivalentPrice /
                 data.veg.avg_form.yield$Yield)
  colnames(data.veg.avg_form.avg_price_yield) = c('Form','Price')
  
  ggplot(data=data.veg.avg_form.avg_price_yield, 
         aes(x=reorder(Form,-Price), y=Price)) + 
    geom_bar(stat='identity', fill='salmon') + theme_minimal() +
    xlab('Vegetable Form') + ylab('Price') +
    geom_text(aes(label=sprintf('%.3f', Price)), vjust=1.6) +
    ggtitle('Average Price per Edible Cup of Vegetable by Form')
  ggsave('visuals/veg_price_per_yield.jpg')
```


## Results

Cheapest form of vegetable by edible cup is canned. The most expensive form is frozen. The yield of dried is very high compard to the others but it becomes equalized when price per cup is involved. Once again, this does not consider calorie count!



# Is a Tomato a fruit?

Determine if the natural clusters based on retail price, yield, and price per cup of fruits and vegetables is reflective of if they are a fruit or vegetable.

## Methodology

```{r kmeans_split_data}
  set.seed(2024)
  
  features = c('RetailPrice', 'Yield', 'CupEquivalentPrice')
  training = select(data, all_of(features))
```


```{r kmeans_training}
  set.seed(2024)

  # create the model
  model = kmeans(training, centers=2)
  
  # grab the model labels for the data
  data['cluster'] = model$cluster
  
  # kmeans labels = {1,2}. We want {0,1}
  data$cluster[data$cluster == 2] = 0
  
  confusion_matrix = table(data$isFruit, data$cluster)
  confusionMatrix(confusion_matrix)
```


```{r kmeans_3d_scatter, message=FALSE, warning=FALSE}
  fig = plot_ly(data=data, x=~RetailPrice, y=~Yield, z=~CupEquivalentPrice,
                type='scatter3d', mode = 'markers', symbol=~I(isFruit), color=~cluster)
  fig = layout(fig, title='Clusters of Fruits and Vegetables Based on 
         Price, Yield, and Price per Cup')
  fig
```


```{r tomato}
  data[grep('Tomato', data$Food),c('Food', 'isFruit', 'cluster')]
```

## Results

We got an accuracy of `0.671`; however, the p-value was `1` so we cannot say we could accurately determine if a product was a fruit or vegetable. Interestingly enough, KMeans determined that grape and cherry tomatoes are fruits but any other type of tomato is a vegetable. I guess that settles the age old question: "is a tomato a fruit?".


# Regression of Fruits and Vegetables

Create a logistic regression model to predict whether an item is a fruit or a vegetable

**Input features:**
`Yield`, `CupEquivalentSize`, `CupEquivalentPrice`

**Output:**
`Vegetable` or `Fruit`

## Methodology

```{r train_test_split}
  # create the train/test split

  set.seed(2024)

  # 10 fruit and 15 veg
  num_fruit = 10
  num_veg = 15
  
  # generate the indicies for the test
  test.fruit.ind = sample(1:length(data.fruit$Food), num_fruit)
  test.veg.ind = sample(1:length(data.veg$Food), num_veg)
  
  # grab the test items from each dataframe
  test.fruit = data.fruit[test.fruit.ind,]
  test.veg = data.veg[test.veg.ind,]
  
  # gather the remaining items
  train.fruit = data.fruit[-test.fruit.ind,]
  train.veg = data.veg[-test.veg.ind,]
  
  # combine the fruit and vegetable test/train dataframes
  test = data.frame(rbind(test.fruit, test.veg))
  train = data.frame(rbind(train.fruit, train.veg))
  
  # shuffle the sets
  test = test[sample(1:nrow(test)),]
  train = train[sample(1:nrow(train)),]
  
  # check to make sure they are shuffled
  head(train)
```

Shuffling the subsets is redundant because there is no index feature of the dataset...

```{r model_creation}
  # training the logistic model
  log_model = glm(isFruit~Yield+CupEquivalentSize+CupEquivalentPrice,
                  family = binomial, data = train)

  summary(log_model)
```

```{r model_testing}
  # testing the model
  log_model.raw_prediction = predict(log_model, test, type="response")
  log_model.prediction = rep(0, nrow(test))
  log_model.prediction[log_model.raw_prediction > 0.5] = 1
  
  # confusion matrix
  log_model.classification_table = table(log_model.prediction, test$isFruit)
  log_model.cm = confusionMatrix(log_model.classification_table)
  
  log_model.cm
```

## Results

Using a 85/15 train/test split of the data, the logistic regression model had a prediction rate of `0.68`. However, the p-value is `0.2735`; therefore, we cannot confidently say that we can predict whether a product is a fruit of vegetable based on retail price, yield, and price per cup.
